---
title: "Spatial regression"
author: "Anusha Bishop"
output: 
  github_document:
    toc: true
    toc_depth: 2
editor_options: 
  markdown: 
    wrap: 72
---

```{r, warning=FALSE, message=FALSE}
#devtools::install_github("AnushaPB/wingen", ref = "release/2.2.0")
library(wingen)

# Spatial packages
library(raster)
library(terra)
library(sf)

# Spatial modelling packages
library(nlme)
library(spatialreg)
library(gstat)
library(spdep)

# Other general packages for data manipulation and visualization
library(tidyverse)
library(cowplot)
library(ggpubr)

load_middle_earth_ex()
```

This tutorial will walk through how to fit spatial regression models to genetic diversity data using example data from the wingen package. This simulated dataset is described in the wingen vignett as such:

"...we will use a subset of the data from the [Bishop et al. (2023)](http://doi.org/10.1111/2041-210X.14090) simulation example. These simulations were created using Geonomics ([Terasaki Hart et al., 2022](https://doi.org/10.1093/molbev/msab175)) to generate a realistic landscape genomic dataset. In this simulation, spatial variation in genetic diversity is produced by varying population size and gene flow across the landscape via heterogeneous carrying capacity and conductance surfaces. These surfaces are based on an example digital elevation model of Tolkien's Middle Earth produced by the Center for Geospatial Analysis at William & Mary ([Robert, 2020](https://scholarworks.wm.edu/asoer/3/))."

# 1. Run wingen 

We will start by running the wingen functions to create a moving window map of genetic diversity. Our predictor of genetic diversity is the raster layer (`lotr_lyr`) that controlled carrying capacity and connectivity in the simulation. In this case, we know that this layer is the primary driver of genetic diversity. 

```{r}
# Create layer for wingen 
lotr_lyr <- rast(lotr_lyr)
names(lotr_lyr) <- "predictor"
lyr <- aggregate(lotr_lyr, 3)

# Run moving window to estimate genetic diversity based on heterozygosity
wgd <- window_gd(lotr_vcf, lotr_coords, lyr, stat = "Ho", wdim = 5)

# Krige results (here I am kriging to the same resolution as the input)
kgd <- winkrige_gd(wgd)

# Extract values at coordinates for model
# Note: I only use the values at coordinates, not the full raster, because I don't want to artifically inflate my sample size
# And I want high confidence in the values I am inputting
df <-
  lotr_coords %>%
  mutate(Ho = terra::extract(kgd, lotr_coords, ID = FALSE)[,1]) %>%
  mutate(predictor = terra::extract(lotr_lyr, lotr_coords, ID = FALSE)[,1]) %>%
  st_as_sf(coords = c("x", "y"), crs = st_crs(lotr_coords)) %>% # Convert to sf object
  bind_cols(lotr_coords) # Add back x/y coordinates which will be used for modelling later

# Model formula: 
f <- formula("Ho ~ predictor")
```

# 2. Fit spatial regression models

## 2.1 Option 1: Variogram and GLS models

Our first option for fitting a spatial model is to use a Generalized Least Squares (GLS) approach with spatially correlated errors. We start by fitting an ordinary least squares (OLS) model and checking the residuals for spatial autocorrelation. We then fit a variogram to these residuals, which quantifies how residual similarity decreases with distance. We then fit different variogram models (exponential, spherical, Gaussian) to estimate the spatial covariance structure and incorporate this structure into a GLS model.

```{r, fig.width=4.5, fig.height=4}
# Fit an OLS model to the data
ols_model <- lm(f, data = df)

# Extract the residuals
df$ols_resids <- residuals(ols_model)

# We can plot the residuals to check for spatial autocorrelation
plt_ols <-
  ggplot(df) +
  geom_point(aes(x = x, y = y, color = ols_resids), cex = 2) +
  scale_color_viridis_c(option = "turbo") +
  labs(title = "OLS Residuals", col = "Residuals")
plt_ols
# We can see spatial clustering in the residuals, indicating spatial autocorrelation
# We can also perform a Moran's I test to quantify this, but I will do that later

# Fit a variogram to the residuals
ols_variogram <- variogram(ols_resids ~ 1, data = df)

# Test different variogram models
exp_var <- fit.variogram(ols_variogram, model = vgm("Exp"))
sph_var <- fit.variogram(ols_variogram, model = vgm("Sph"))
gau_var <- fit.variogram(ols_variogram, model = vgm("Gau"))

# Plot the variograms
plot(ols_variogram, exp_var, main = "OLS Residuals Variogram (Exponential)")
plot(ols_variogram, sph_var, main = "OLS Residuals Variogram (Spherical)")
plot(ols_variogram, gau_var, main = "OLS Residuals Variogram (Gaussian)")
# Based on visual inspection, the models all seem to fit well
```

```{r, fig.width = 9, fig.height = 4}
# We can fit GLS models using the variogram models and compare them based on AIC
# exp_model <- gls(f, correlation = corExp(form = ~ x + y), data = df) # False convergence error
sph_model <- gls(f, correlation = corSpher(form = ~ x + y), data = df)
gau_model <- gls(f, correlation = corGaus(form = ~ x + y), data = df)

# The AIC values are similar, but the best is the spherical model
AIC(sph_model, gau_model)

# We can plot the residuals of this model to confirm that they are now spatially uncorrelated
# Note: we most use type = "normalized" to get standardized residuals, otherwise the residuals will still show spatial autocorrelation
df$gls_resids <- residuals(sph_model, type = "normalized") 

plt_gls <-
  ggplot(df) +
  geom_point(aes(x = x, y = y, color = gls_resids), cex = 2) +
  scale_color_viridis_c(option = "turbo") +
  labs(title = "GLS Residuals", col = "Residuals")
plot_grid(plt_ols, plt_gls, nrow = 1)
# We can see that the residuals are now spatially uncorrelated
```

A note about nuggets:
By default, the GLS models do not fit a nugget term. This means that they assume residuals at the same location are perfectly correlated, which may not always be the case in real-world data. If you suspect that there is unstructured microscale variance (random noise at zero distance), you can include a nugget term in your model by setting `nugget = TRUE` in the correlation structure. However, this can absorb spatial signal and cause predictor effects to appear insignificant. Because of this, I fit my models without a nugget term, because I don't expect to have much microscale variation in heterozygosity.
```{r}
no_nugget <- gls(f, correlation = corSpher(form = ~ x + y, nugget = FALSE), data = df)
nugget <- gls(f, correlation = corSpher(form = ~ x + y, nugget = TRUE), data = df)

# In this case, there is no major difference between the nugget vs no nugget models.
summary(no_nugget)
summary(nugget)

AIC(no_nugget, nugget)
```

## 2.2 Option 2: Spatial error and lag models

Our second option for fitting a spatial model is to model spatial dependence explicitly using spatial lag and spatial error models. These models extend ordinary least squares (OLS) by incorporating spatial relationships between observations, either in the response variable (spatial lag; SLM) or in the error structure (spatial error; SEM). The trickiest part of this approach is to create a spatial weights matrix defining your neighborhoods. This can be done in a few different ways, but the most common is to use k-nearest neighbors or distance-based neighbors. Most tutorials on using SEM/SLM models use examples where the data are adjacent polygons (e.g., counties or census tracts), which makes neighbors more intuitive. In comparison, picking the right approach for continuous data distributions is challenging; what I have ended up doing is fitting a variogram to the residuals of an OLS model, and then using the range of that variogram to create a distance-based neighbor list.

https://chrismgentry.github.io/Spatial-Regression/

First, we construct our neighborhood weights matrix using the range from the variogram we fit above:
```{r}
# We have already fit a variogram above, so I am going to use those results and extract the range from the spherical model
vgm_range <- sph_var %>% filter(model == "Sph") %>% pull(range)

# The range is the distance at which the spatial autocorrelation between points becomes negligible. On a variogram, this is the distance where the variogram levels off.
print(vgm_range)
plot(ols_variogram, sph_var, main = "OLS Residuals Variogram (Spherical)") # Levels off around 40

# Create a spatial weights matrix using the range from the variogram
nb <- dnearneigh(df, d1 = 0, d2 = vgm_range)
listw <- nb2listw(nb, style = "W", zero.policy = TRUE) # zero.policy = TRUE allows for zero weights, which is necessary if some points have no neighbors within the specified distance

# Plot the neighbors to visualize the spatial weights matrix
plot(nb, st_coordinates(df), main = paste0("Neighborhoods (Range = ", round(vgm_range, 2), ")"))
```

 We can use this neighborhood weights matrix to perform Lagrange Multiplier tests for spatial dependence:

| Test       | Tests for                                | Null hypothesis             |
| ---------- | ---------------------------------------- | --------------------------- |
| `SARMA`    | Any spatial dependence (lag or error)    | No spatial dependence       |
| `RSlag`    | Spatial lag dependence                   | No spatial lag dependence   |
| `RSerr`    | Spatial error dependence                 | No spatial error dependence |
| `adjRSlag` | Spatial lag dependence (robust to error) | No spatial lag dependence   |
| `adjRSerr` | Spatial error dependence (robust to lag) | No spatial error dependence |

```{r}
# Run the tests for spatial dependence
lm.RStests(ols_model, listw, test = "all")

# Results:
# SARMA = 84.627, df = 2, p-value < 2.2e-16
# adjRSlag = 2.0598, df = 1, p-value = 0.1512
# adjRSerr = 11.149, df = 1, p-value = 0.0008408
# RSlag = 73.479, df = 1, p-value < 2.2e-16
# RSerr = 82.568, df = 1, p-value < 2.2e-16
```

Based on these results, we should probably fit a Spatial Error Model (SEM) because we have significant adjRSerr, but non-significant adjRSlag (i.e., no strong evidence for lag dependence when adjusting for error dependence.)

```{r, fig.height=4, fig.width=14}
# Fit an SEM model
sem_model <- errorsarlm(f, data = df, listw = listw, zero.policy = TRUE)
summary(sem_model)

# We can also fit an SLM model, for comparison
slm_model <- lagsarlm(f, data = df, listw = listw, zero.policy = TRUE)
summary(slm_model)

# Extract residuals
# Note: here we don't need to use type = "normalized" because the residuals are already standardized by the model
df$sem_resids <- residuals(sem_model)
df$slm_resids <- residuals(slm_model)
df$sem_resids <- df$Ho - fitted(sem_model)
df$slm_resids <- df$Ho - fitted(slm_model)

# We can compare the models based on AIC
AIC(ols_model, sem_model, slm_model)
# Both spatial models are better than OLS, but SEM is better than SLM

# We can also compare Moran's I of the residuals to confirm that the SEM model has reduced spatial autocorrelation
moran.test(df$ols_resids, listw, zero.policy = TRUE) # Moran's I = 0.21 (p < 2.2e-16)
moran.test(df$slm_resids, listw, zero.policy = TRUE) # Moran's I = 0.04 (p = 0.005)
moran.test(df$sem_resids, listw, zero.policy = TRUE) # Moran's I = 0.04 (p = 0.003)
# Both spatial models reduce Moran's I in the residuals, in comparison to the OLS model

# We can plot the residuals of the SEM model to confirm that they are now spatially uncorrelated
plt_sem <- 
  ggplot(df) +
  geom_point(aes(x = x, y = y, color = sem_resids), cex = 2) +
  scale_color_viridis_c(option = "turbo") +
  labs(title = "SEM Residuals", col = "Residuals")

plt_slm <-
  ggplot(df) +
  geom_point(aes(x = x, y = y, color = slm_resids), cex = 2) +
  scale_color_viridis_c(option = "turbo") +
  labs(title = "SLM Residuals", col = "Residuals") 

plot_grid(plt_ols, plt_sem, plt_slm, nrow = 1)
# The SLM/SEM residuals still show spatial autocorrelation
```

We can then compare our SEM/SLM models with the GLS model:
```{r, fig.height = 4, fig.width = 19}
# You will get a warning that the models are not all fitted to the same number of observations, but I believe this is just an artifact of the different ways the models are fit
AIC(ols_model, sph_model, sem_model, slm_model)
# The spherical model is the best fit here

# We can plot the residuals and confirm
plot_grid(plt_ols, plt_gls, plt_sem, plt_slm, nrow = 1)
```

In this example, the spherical GLS model is the best fit, but I have also had cases where the SEM/SLM models are better. It really depends on the data and the spatial structure of the residuals. In this example, there is only one driver of genetic diversity, the habitat suitability/connectivity layer (i.e., `lotr_lyr`), so our predictor is the only variable that is driving genetic diversity.

# 3. Plot results

## 3.1 Predicted vs. observed plots

The first plot I like to make is a scatterplot of the predicted values vs. the observed values. This helps us see how well the model fits the data and whether there are any patterns in the residuals.

Below we plot the predicted values from the GLS model against the observed heterozygosity values. 

```{r}
df$gls_predict <- predict(sph_model)

ggplot(df, aes(x = gls_predict, y = Ho)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Predicted Heterozygosity", y = "Observed Heterozygosity") +
  theme_classic() +
  stat_cor()
```

If we do the same thing for the SEM model, we see that the plot of observed versus predicted values is quite different. 

```{r}
df$sem_predict <- predict(sem_model) %>% data.frame() %>% pull(fit)

ggplot(df, aes(x = sem_predict, y = Ho)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Predicted Heterozygosity", y = "Observed Heterozygosity") +
  theme_classic() +
  stat_cor()
```

This is because, in the GLS model, predictions are based only on the fixed effects (spatial autocorrelation is used to estimate the coefficients but not to generate predictions). In contrast, the SEM model incorporates the spatial effect into its predictions, so the predicted values account for the neighboring structure.

If you plot the predicted values for the SEM model from the original data (i.e, treat the original data like it is "new" data), then you will see the predictions are similar to the GLS model because it excludes the spatially structured error term and only uses the fixed effects.

```{r}
df$sem_predict_new <- predict(sem_model, newdata = df) %>% data.frame() %>% pull(fit)

ggplot(df, aes(x = sem_predict_new, y = Ho)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Predicted Heterozygosity", y = "Observed Heterozygosity") +
  theme_classic() +
  stat_cor()
```

In terms of which predicted versus observed plot is better for the SEM model, it depends on the goal: if you want to assess the in-sample fit, use the first `predict()` values because they include the spatial error component; if you want to evaluate the model’s predictive performance on new data, use the `predict(newdata)` predicted values since they exclude the spatial error and reflect how the model generalizes.

## 3.2 Predictor plots 

We can plot the predictor against the response variable to see how well each predictor explains the variation in the response.

```{r}
ggplot(df, aes(x = predictor, y = Ho)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Predictor", y = "Observed Heterozygosity") +
  theme_classic() +
  stat_cor()
```

The problem with this plot is that it does not account for the spatial autocorrelation in the data, so it may overestimate the strength of the relationship between the predictor and the response. 

If you have multiple predictors, you can use component-plus-residual (C+R) plots, also known as partial residual plots, to visualize the marginal relationship between each predictor and the response variable while controlling for the other predictors. These plots work by adding the estimated contribution of a predictor back to the model residuals, allowing you to see the predictor’s effect after adjusting for the effects of all other variables.

Note: As far as I am aware, you should not use partial regression plots to assess spatial regression models. Partial regression plots show the residuals of the response variable regressed on all other predictors plotted against the residuals of the predictor regressed on the other predictors. These plots are made for OLS models where the residuals are assumed to be independent.

Other resources:

- https://www.paulamoraga.com/book-spatial/index.html

- https://chrismgentry.github.io/Spatial-Regression/#1_Introduction_to_Spatial_Regression

- https://bookdown.org/hhwagner1/LandGenCourse_book/

*AI Acknowledgement: portions of this vignette include text and examples generated with the assistance of ChatGPT (OpenAI)*